---
title: "Triming"
output: html_document
date: "2025-03-05"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      #send figures generated in this folder
                      fig.path = "../figures/02_AssignASVs/")
```


# Set up the environment
## Set seed
```{r set-seed}
set.seed{20020206}
```


## Load packages
```{r load-packages}
pacman::p_load(tidyverse, devtools, dada2,
               patchwork, DT, install = FALSE)

```


#Goals

1. Infer Errors in our sequences, sepearately on forward and reverse reads.
2. Assign ASVs on both forward and reverse reads separately. Apply the error model.
3. Merge forward and reverse ASVs in "Contiguous ASVs"
4. Generate first draft of ASV count table.
5. Quality Trimming ASV lengths.
6. Remove chimeras.
7. Assign Taxonomy with Silva Database.
8. Write out relevant files: 'asv_table', 'asv_fasta', 'tax_table', and 'sample_data'.


## Input

1. Filtered fastq files generated from '01_QualityTrimming.Rmd'.
2. Sample Name vector.


## Output

1. 'asv_table'
2. 'asv_fasta'
3. 'tax_table'
4. 'sample_data'



# Load filtered fastq files
```{r load-filtered-files}
# place filtered seq filed into a variable
filtered_fastqs_path <- "data/01_DADA2/02_filtered_fastqs"

# Intuition check
filtered_fastqs_path

# Create forward vector
filtered_forward_reads <-
  list.files(filtered_fastqs_path, pattern = "R1_filtered.fastq.gz",
             full.names = TRUE)

# Check
filtered_forward_reads[1:5]

# Create forward vector
filtered_reverse_reads <-
  list.files(filtered_fastqs_path, pattern = "R2_filtered.fastq.gz",
             full.names = TRUE)

# Check
filtered_reverse_reads[1:5]
```


# Sample Names
```{r sample-names}
# Create vector of sample names from the filenames 
sample_names <- sapply(strsplit(basename(filtered_forwards_reads), "_"), `[`,1) 

# Intuition Check 
head(sample_names)

```


```{r learn-errors}
# Forward Reads
error_forward_reads <-
  learnErrors(filtered_forward_reads, multithread = 6)

# Plot
forward_error_plot <-
  plotErrors(error_forward_reads, nominalQ = TRUE) +
  labs(title = "Forward Read: Error Model")
forward_error_plot

# Reverse Reads
error_reverse_reads <-
  learnErrors(filtered_reverse_reads, multithread = 6)

# Plot
reverse_error_plot <-
  plotErrors(error_reverse_reads, nominalQ = TRUE) +
  labs(title = "Reverse Read: Error Model")
reverse_error_plot

# Look at the plots together
forward_error_plot + reverse_error_plot
```

# Infer ASVs
```{r infer-ASVs}
# Forward ASVs
dada_forward <-
  dada(filtered_forward_reads,
       err = error_forward_reads,
       multithread = 6)

# Take a look at the data
typeof(dada_forward)
dada_forward

# Reverse ASVs
dada_reverse <-
  dada(filtered_reverse_reads,
       err = error_reverse_reads,
       multithread = 6)

# Take a look at the data
typeof(dada_reverse)
dada_reverse[30]

```


# Merge Forward and Reverse ASVs
```{r merge-ASVs}
merged_ASVs <-
  mergePairs(dada_forward, filtered_forward_reads,
             dada_reverse, filtered_reverse_reads,
             verbose = TRUE)


# Evaluate the data output
typeof(merged_ASVs)
length(merged_ASVs)
names(merged_ASVs)

# Inspect further
head(merged_ASVs)
merged_ASVs$`022um-Control_R1_filtered.fastq.gz`


```


# Create a rough
```{r raw-ASV-count-table}
# Raw ASV
raw_ASV_table <- makeSequenceTable(merged_ASVs)

# Check
dim(raw_ASV_table)
typeof(raw_ASV_table)
class(raw_ASV_table)

# Write out raw_ASV_table
write.table(raw_ASV_table, file = "data/01_DADA2/raw_ASV_counts.tsv",
            sep = "\t", quote = FALSE, col.names = NA)


```


# Assess the ASV Length
```{r assess-ASV-length}
# Creating a table to inspect the distribution of ASV lengths
table(nchar(getSequences(raw_ASV_table)))

#Plot
data.frame(ASV_Length = nchar(getSequences(raw_ASV_table))) %>%
  ggplot(aes(x = ASV_Length)) + 
  geom_histogram() +
  scale_x_continuous(limits = c(0, 500)) +
  labs(title = "Raw ASV Lengths",
       y = "Number of ASVs", x = "ASV Sequence Length (bps)")
```


# Trim ASVs
```{r trim-ASVs}

# Only pull ASVs that have a length of 245 bps
raw_ASV_table_trimmed <-
  raw_ASV_table[,nchar(getSequences(raw_ASV_table)) == 245]
raw_ASV_table_trimmed

# Intuition Check
table(nchar(getSequences(raw_ASV_table_trimmed)))

#Plot
data.frame(ASV_Length = nchar(getSequences(raw_ASV_table_trimmed))) %>%
  ggplot(aes(x = ASV_Length)) + 
  geom_histogram() +
  scale_x_continuous(limits = c(0, 500)) +
  labs(title = "Trimed Raw ASV Lengths",
       y = "Number of ASVs", x = "ASV Sequence Length (bps)")
```


# Remove Chimeras
```{r remove-chimeras}
noChimeras_ASV_table <-
  removeBimeraDenovo(raw_ASV_table_trimmed,
                     method = "consensus",
                     multithread = 6,
                     verbose = TRUE)

# Structure of data
dim(noChimeras_ASV_table)
dim(raw_ASV_table_trimmed)

```


# Track the number of reads DARA2 workflow

# Assign Taxonomy












# Session Information
```{r session-info}
# Ensure reproducibility with package version information
devtools::as.package()

```
